The Implications of Local Fluctuations in the Galactic Midplane for Dynamical Analysis in the \textit{Gaia} Era

### 2019-07-23 paper/ms.tex
Update title

\title{The Implications of Disk Oscillations for Dynamical Analysis in the
\textit{Gaia} Era}

@@

Orbital properties of stars, computed from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to understand the
structure and evolution of the Galaxy. Stellar actions, computed from orbits,
have the attractive quality of being invariant under certain assumptions and
are therefore used as quantitative labels of a star's orbit. We report a
subtle but important systematic error that is induced in the actions as a
consequence of \textbf{disk oscillations} or local midplane variations expected for the Milky Way. This
error is difficult to model because it is non-Gaussian and bimodal, with
neither mode peaking on the null value. An offset in the vertical position of
the Galactic midplane of $\sim15\,\pc$ for a thin disk-like orbit or $\sim
120\,\pc$ for a thick disk-like orbit induces a $25\%$ systematic error in the
vertical action $J_z$. In FIRE simulations of Milky Way-mass galaxies, these
variations are on the order of $\sim100\,\pc$ at the solar circle. From
observations of the mean vertical velocity variation of
$\sim5\textup{--}10\,\kms$ with radius, we estimate that the Milky Way
midplane variations are $\sim60\textup{--}170\,\pc$, consistent with
three-dimensional dust maps. Action calculations and orbit integrations, which
assume the global and local midplanes are identical, are likely to include
this induced error, depending on the volume considered. Variation in the local
standard of rest or distance to the Galactic center causes similar issues. The
variation of the midplane must be taken into account when performing dynamical
analysis across the large regions of the disk accessible to \textit{Gaia} and
future missions.

>>>>

\title{The Implications of Local Fluctuations in the Galactic Midplane for
Dynamical Analysis in the \textit{Gaia} Era}

@@

Orbital properties of stars, computed from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to understand the
structure and evolution of the Galaxy. Stellar actions, computed from orbits,
have the attractive quality of being invariant under certain assumptions and
are therefore used as quantitative labels of a star's orbit. We report a
subtle but important systematic error that is induced in the actions as a
consequence of local midplane variations expected for the Milky Way. This
error is difficult to model because it is non-Gaussian and bimodal, with
neither mode peaking on the null value. An offset in the vertical position of
the Galactic midplane of $\sim15\,\pc$ for a thin disk-like orbit or $\sim
120\,\pc$ for a thick disk-like orbit induces a $25\%$ systematic error in the
vertical action $J_z$. In FIRE simulations of Milky Way-mass galaxies, these
variations are on the order of $\sim100\,\pc$ at the solar circle. From
observations of the mean vertical velocity variation of
$\sim5\textup{--}10\,\kms$ with radius, we estimate that the Milky Way
midplane variations are $\sim60\textup{--}170\,\pc$, consistent with
three-dimensional dust maps. Action calculations and orbit integrations, which
assume the global and local midplanes are identical, are likely to include
this induced error, depending on the volume considered. Variation in the local
standard of rest or distance to the Galactic center causes similar issues. The
variation of the midplane must be taken into account when performing dynamical
analysis across the large regions of the disk accessible to \textit{Gaia} and
future missions.
